---
title: "data_prep"
output: html_document
---

```{r setup}
library(plyr)
library(dplyr)
library(Seurat)
library(patchwork)
library(EpicTools)
library(ggpubr)
library(sva)
library(RUVSeq)
library(tidyverse)
library(SingleR)
library(SingleCellExperiment)
library(scRNAseq)
library(scater)
```



```{r}
suppressPackageStartupMessages({
  library(ggplot2)
  library(devtools)
})
```


Type of work flow 
```{r}
if_do_combat = FALSE
```

Loading list of genes ones that were differently expressed based on supplementary materials from the paper. 
```{r }
gene_names <- read.csv(file = 'genes.csv',  header = FALSE)
gene_names <- unique(gene_names[,1])
```
Loading data
```{r }
path = "data/"
cm.list = paste0(path, list.files(pattern = "*.matrices.rds", path = path))

cm.files <- lapply(cm.list, readRDS)
names(cm.files) <- sub(path,"",
                       sub("\\_cell.counts.matrices.rds", "", cm.list))
```

Pre-processing
```{r }
cm.pp <- mapply(EpicPreHS, cm.files, orig.ident = names(cm.files), SIMPLIFY = F)
covid_combined.emat <- mergeCM(cm.pp, type = "emat")
```
Limiting genes to ones that were differently expressed based on supplementary materials from the paper. 
```{r }
r<- row.names(covid_combined.emat)
b  <-gene_names %in% r
covid_combined.emat <- covid_combined.emat[gene_names[b],]
```
Checking for missing data
```{r }
rows_missing <- apply(covid_combined.emat, 1, function(x) any(is.na(x)))
sum(rows_missing)
```
No missing data was found

Creating batches
```{r }
if (if_do_combat){
  col <- colnames(covid_combined.emat)
  col_split <- sapply(col, strsplit, split = ".",  fixed=TRUE)
  df <- ldply (col_split, data.frame)
  df <- df[ c(TRUE,FALSE), ]
  covid_combined.batch <- list(df[,2], df[,1])
  names(covid_combined.batch) <- c("patient", "cell")
}

```



Creating combat object
```{r }
if (if_do_combat){
  mod0 <- model.matrix(~ 1, data = t(covid_combined.emat))
  covid_combined.combat <- ComBat(
      dat = covid_combined.emat, 
      batch = factor(covid_combined.batch$patient), 
      mod = mod0,
      par.prior = TRUE,
      prior.plots = TRUE
  )
}
```
Creating Seurat object
```{r}
if (!if_do_combat){
  covid_combined <- CreateSeuratObject(counts = covid_combined.emat, min.cells = 10, names.field = 1, names.delim = "\\.")
}

if (if_do_combat){
  covid_combined <- CreateSeuratObject(counts = covid_combined.combat, min.cells = 10, names.field = 1, names.delim = "\\.")
}
```
Adding metadata
```{r }
covid_metadata <- read_csv("meta_data.csv")
seurat_metadata <- covid_combined@meta.data
seurat_metadata <- seurat_metadata[,!(colnames(seurat_metadata)) %in% 
                                     setdiff(colnames(covid_metadata), "orig.ident")]
metadata_combined <- merge(seurat_metadata, covid_metadata, by = "orig.ident")
rownames(metadata_combined) <- rownames(covid_combined@meta.data)
covid_combined@meta.data <- metadata_combined
```

```{r}
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^MT-", col.name = "percent.mt")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RPS", col.name = "percent.rps")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RPL", col.name = "percent.rpl")
covid_combined <- PercentageFeatureSet(covid_combined, pattern = "^RNA\\d8S5", col.name = "percent.rrna")
```

Saving VlnPlot for quality check
```{r }
png("plots/vln_org.png",height=500,width=900)
VlnPlot(covid_combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0.0)
dev.off()
png("plots/vln_prg_2.png",height=500,width=900)
VlnPlot(covid_combined, features = c( "percent.rps", "percent.rpl", "percent.rrna"), ncol = 3, pt.size = 0.0)
dev.off()
```

Additional filtering
```{r}
if (!if_do_combat){
  covid_combined <- subset(covid_combined, subset =  nCount_RNA < 10000 & percent.mt < 18 & percent.mt > 0 & nFeature_RNA < 1200 )
}

if (if_do_combat){
  covid_combined <- subset(covid_combined, subset =  nCount_RNA < 10000 & percent.mt < 18 & percent.mt > 0 )
}
```
Save to file for azimuth
```{r }
saveRDS(covid_combined, file = "seurat/covid_org.rds")
```

Normalize and scale
```{r }
covid_combined <- NormalizeData(covid_combined)
covid_combined <- FindVariableFeatures(covid_combined, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(covid_combined)
covid_combined <- ScaleData(covid_combined, features = all.genes)
```
Clustering
```{r }
covid_combined <- RunPCA(covid_combined, verbose = FALSE)
covid_combined <- RunUMAP(covid_combined, dims = 1:50, verbose = FALSE)
covid_combined <- FindNeighbors(covid_combined, dims = 1:50, verbose = FALSE)
covid_combined <- FindClusters(covid_combined, resolution = 1, verbose = FALSE)
```
Checking cluster quality
```{r}
covid_combined.markers <- FindAllMarkers(covid_combined, only.pos = TRUE, min.pct = 0.25)
covid_combined.markers %>% group_by(cluster)  %>% top_n(n = 10, wt = avg_log2FC)

```
Saving cluster quality check
```{r }
write.csv(covid_combined.markers,"tables/cluster_org_quality.csv", row.names = FALSE)
```

SingleR annotation
```{r cell annotation}
ref <- HumanPrimaryCellAtlasData()
common <- intersect(rownames(ref), rownames(covid_combined.emat))
combined.emat.singler <- covid_combined.emat[common,]
ref <- ref[common,]
combined.emat.singler.sce <- SingleCellExperiment(assays = list(counts = combined.emat.singler))
combined.emat.singler.sce <- logNormCounts(combined.emat.singler.sce)
```

Predicting cell type
```{r }
singler.pred <- SingleR(test = combined.emat.singler.sce@assays@data@listData$logcounts, ref = ref, labels = ref$label.main)
singler.pred <- singler.pred[colnames(covid_combined),]
table(singler.pred$labels)
```
Saving cell type to file
```{r }
write.csv(singler.pred[,c('labels','pruned.labels')],"tables/seurat_combat_cell_ad.csv")
```

Heatmap of cell types
```{r }
png("plots/heatmap_org.png",height=500,width=900)
plotScoreHeatmap(singler.pred)
dev.off()
```

Annotating clusters witjh cell types
```{r }
singler.results <- merge(data.frame(cell = rownames(singler.pred), singler = singler.pred$labels), 
                         data.frame(cell = rownames(covid_combined@meta.data), 
                                    cluster = covid_combined@meta.data$seurat_clusters), 
                         by = "cell", 
                         all.y = FALSE)
singler.results$cell <- NULL
singler.results$count <- 1
singler.results <- aggregate(count ~ ., singler.results, FUN = sum)
singler.final <- singler.results %>% group_by(cluster) %>% top_n(n = 1, wt = count)
singler.final
covid_combined$singler <- singler.pred$labels
```

Saving clusters annotations to a file
```{r}
write.csv(singler.results,"tables/seurat_combat_cluster_ad.csv", row.names = FALSE)
```


Adding cell types to Seurat object
```{r}
covid_combined$cell.type.fine  <- 'NULL'
for (row in 1:nrow(singler.final)){
  covid_combined$cell.type.fine[covid_combined$seurat_clusters == singler.final[row,]$cluster ] <- singler.final[row,]$singler
}

```

Saving Seurat object for further analyze
```{r}
saveRDS(covid_combined, file = "seurat/covid_combat_seurat3.rds")
```
